---
title: "Lab 3: DML Implementation with DoubleML"
subtitle: "Code-Along Lab"
author: "Sean Sylvia"
date: "March 9, 2026"
slug: "lab-3-dml"
categories: [Lab, Code-Along, DML, DoubleML]
description: "Hands-on implementation of Double Machine Learning using the DoubleML R package. Fit PLR and IRM models, compare learners, and apply DML to real data."
format:
  html:
    toc: true
    toc-depth: 2
    code-tools: true
editor: visual
execute:
  eval: false
---

## Accept the Assignment

[**Click here to accept Lab 3 on GitHub Classroom**](https://classroom.github.com/a/cTEpnvFg)

This will create your personal repository for this lab.

**To get started:**

1. Accept the assignment using the link above
2. Clone your repository locally: `git clone <your-repo-url>`
3. Open `hpm883-lab.Rproj` in RStudio or Positron
4. Run `renv::restore()` to install packages
5. Open `lab-3-dml.qmd` and follow along

::: callout-important
## Code-Along Session (Remote, Zoom)

All code chunks have `eval: false` — run them yourself in RStudio or Positron. This is a **75-minute guided session**, not a rendering exercise.
:::

## Overview

In Session 2.3, we derived the DML recipe:

```
1. Split data into K folds
2. For each fold k:
   a. Train ML models for g(X) = E[Y|X] and m(X) = E[D|X] on other folds
   b. Predict on fold k to get residuals: Ỹ = Y - ĝ(X) and D̃ = D - m̂(X)
3. Combine all residuals
4. θ̂ = regress Ỹ on D̃
5. Compute SEs from influence function formula
```

Today we implement this in R using the `DoubleML` package, which handles the cross-fitting, nuisance estimation, and inference machinery for us. By the end of this session, you will:

1. Set up `DoubleMLData`, `DoubleMLPLR`, and `DoubleMLIRM` objects
2. Fit DML models with different ML learners (LASSO, Random Forest, XGBoost)
3. Interpret DML output — coefficients, standard errors, confidence intervals
4. Diagnose common issues (overlap, learner choice, fold sensitivity)
5. Apply DML to real data (401(k) participation)

------------------------------------------------------------------------

## Setup & Orientation

### Load Packages

```{r}
#| label: setup
#| message: false
#| warning: false

library(DoubleML)
library(mlr3)
library(mlr3learners)
library(data.table)

# ML backends
library(glmnet)    # LASSO / Ridge / Elastic Net
library(ranger)    # Random Forest
library(xgboost)   # Gradient Boosting

# Visualization
library(ggplot2)

# Reproducibility
set.seed(883)
```

### The DoubleML Architecture

`DoubleML` is built on three layers:

| Layer | Class | What it does |
|-------|-------|-------------|
| **Data** | `DoubleMLData` | Wraps your dataset, specifying Y, D, X roles |
| **Model** | `DoubleMLPLR`, `DoubleMLIRM`, ... | Defines the causal model and moment conditions |
| **Learner** | `mlr3` learners | Specifies which ML algorithm estimates nuisance functions |

The workflow is always: **Data → Model → Fit → Summarize**.

::: callout-tip
## The `mlr3` Connection

`DoubleML` uses the `mlr3` ecosystem for ML learners. You specify learners with `lrn("regr.ranger")` or `lrn("classif.ranger")`. The `mlr3learners` package provides access to `glmnet`, `ranger`, `xgboost`, and many others. You don't need to learn `mlr3` deeply — just the `lrn()` function.
:::

------------------------------------------------------------------------

## Exercise 1: Basic PLR with Simulated Data

### The Partially Linear Model

Recall the PLM:

$$
Y = D\theta_0 + g_0(X) + \varepsilon, \quad E[\varepsilon | D, X] = 0
$$

$$
D = m_0(X) + V, \quad E[V | X] = 0
$$

We want to estimate $\theta_0$ — the causal effect of $D$ on $Y$ — while allowing $g_0$ and $m_0$ to be arbitrary nonlinear functions.

### Step 1: Generate Data with Known Truth

We'll create a DGP where the true $\theta_0 = 0.5$ and the confounding functions $g_0$ and $m_0$ are nonlinear.

```{r}
#| label: dgp-plr

# --- Data Generating Process ---
n <- 2000
p <- 10  # Number of covariates

# Covariates: correlated normals
Sigma <- toeplitz(0.5^(0:(p-1)))
X <- matrix(rnorm(n * p), n, p) %*% chol(Sigma)
colnames(X) <- paste0("X", 1:p)

# Nonlinear confounding functions
g0 <- sin(X[,1]) + X[,2]^2 + X[,3] * X[,4]      # E[Y|X] component
m0 <- 0.5 * X[,1] + 0.25 * X[,2]^2 + X[,3]      # E[D|X] component

# Treatment and outcome
theta0 <- 0.5  # TRUE CAUSAL EFFECT
D <- m0 + rnorm(n)
Y <- D * theta0 + g0 + rnorm(n)

# Combine into data.table
sim_data <- data.table(Y = Y, D = D, X)

cat("True theta_0 =", theta0, "\n")
cat("n =", n, ", p =", p, "\n")
```

::: callout-note
## Why Simulated Data First?

With simulated data, we **know** the true parameter. This lets us check whether our method recovers it. In applied work, you never know the truth — so building intuition with known DGPs is essential.
:::

### Step 2: Create the DoubleMLData Object

```{r}
#| label: dml-data

# Wrap data for DoubleML
dml_data <- DoubleMLData$new(
  data = sim_data,
  y_col = "Y",
  d_cols = "D",
  x_cols = paste0("X", 1:p)
)

# Inspect
print(dml_data)
```

### Step 3: Fit DoubleMLPLR

```{r}
#| label: plr-fit

# Specify learners for nuisance functions
ml_l <- lrn("regr.ranger", num.trees = 500, min.node.size = 5)  # For E[Y|X]
ml_m <- lrn("regr.ranger", num.trees = 500, min.node.size = 5)  # For E[D|X]

# Create and fit the PLR model
dml_plr <- DoubleMLPLR$new(
  data = dml_data,
  ml_l = ml_l,
  ml_m = ml_m,
  n_folds = 5,
  n_rep = 1,
  score = "partialling out"
)

dml_plr$fit()
```

### Step 4: Interpret the Output

```{r}
#| label: plr-summary

# Summary with coefficient, SE, t-stat, p-value, CI
dml_plr$summary()

# Extract key quantities
cat("\nEstimate: ", dml_plr$coef, "\n")
cat("Std Error:", dml_plr$se, "\n")
cat("95% CI:   [", dml_plr$confint()[1], ",", dml_plr$confint()[2], "]\n")
cat("True θ₀:  ", theta0, "\n")
```

::: callout-important
## 🤔 Think About It

Does the 95% confidence interval contain the true $\theta_0 = 0.5$? Run the code a few times (changing the seed). How often does it cover?
:::

### Step 5: The Failure Case — Naive OLS

What happens if we ignore the nonlinear confounding and just run OLS?

```{r}
#| label: naive-ols

# Naive OLS: regress Y on D and all X linearly
ols_fit <- lm(Y ~ D + X, data = data.frame(Y, D, X))

cat("Naive OLS estimate:", coef(ols_fit)["D"], "\n")
cat("DML estimate:      ", dml_plr$coef, "\n")
cat("True theta_0:      ", theta0, "\n")
```

::: callout-warning
## ⚠️ Common Pitfall: Regularization Bias

Naive OLS with many covariates suffers from two problems: (1) it can't capture the nonlinear confounding, and (2) if you use LASSO for variable selection first, the "post-LASSO" estimator inherits regularization bias. DML solves both via orthogonalization + cross-fitting.
:::

------------------------------------------------------------------------

## Exercise 2: Learner Comparison

Does the choice of ML learner for nuisance estimation matter? Let's compare three popular options.

### Three Learners, Same DGP

```{r}
#| label: learner-setup

# Define three learner pairs
learners <- list(
  "LASSO" = list(
    ml_l = lrn("regr.cv_glmnet", s = "lambda.min", alpha = 1),
    ml_m = lrn("regr.cv_glmnet", s = "lambda.min", alpha = 1)
  ),
  "Random Forest" = list(
    ml_l = lrn("regr.ranger", num.trees = 500, min.node.size = 5),
    ml_m = lrn("regr.ranger", num.trees = 500, min.node.size = 5)
  ),
  "XGBoost" = list(
    ml_l = lrn("regr.xgboost", nrounds = 100, max_depth = 3, eta = 0.1,
               verbose = 0),
    ml_m = lrn("regr.xgboost", nrounds = 100, max_depth = 3, eta = 0.1,
               verbose = 0)
  )
)
```

### Fit All Three

```{r}
#| label: learner-comparison

results <- data.frame(
  Learner = character(),
  Estimate = numeric(),
  SE = numeric(),
  CI_lower = numeric(),
  CI_upper = numeric(),
  stringsAsFactors = FALSE
)

for (name in names(learners)) {
  dml_fit <- DoubleMLPLR$new(
    data = dml_data,
    ml_l = learners[[name]]$ml_l,
    ml_m = learners[[name]]$ml_m,
    n_folds = 5,
    n_rep = 1,
    score = "partialling out"
  )
  dml_fit$fit()

  results <- rbind(results, data.frame(
    Learner = name,
    Estimate = dml_fit$coef,
    SE = dml_fit$se,
    CI_lower = dml_fit$confint()[1],
    CI_upper = dml_fit$confint()[2]
  ))
}

print(results)
```

### Forest Plot

```{r}
#| label: forest-plot

ggplot(results, aes(x = Estimate, y = Learner)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = CI_lower, xmax = CI_upper), height = 0.2) +
  geom_vline(xintercept = theta0, linetype = "dashed", color = "red") +
  labs(
    title = "DML Estimates by Nuisance Learner",
    subtitle = paste("True θ₀ =", theta0, "(dashed red line)"),
    x = "Estimated Treatment Effect",
    y = ""
  ) +
  theme_minimal(base_size = 14)
```

::: callout-note
## 🤔 Think About It

- Do all three learners recover the true $\theta_0$?
- Which has the smallest confidence interval? Why?
- When our DGP has nonlinear confounding, which learner should struggle most?

**Key insight:** DML's orthogonality property means the final estimate is **robust to moderate errors in nuisance estimation**. As long as the ML learners converge at $n^{-1/4}$ rate, the treatment effect estimate is $\sqrt{n}$-consistent. This is why all three can work — but LASSO may struggle when the true functions are highly nonlinear.
:::

------------------------------------------------------------------------

## Exercise 3: IRM for Binary Treatment

When treatment $D$ is binary (0/1), we switch from PLR to the **Interactive Regression Model** (IRM):

$$
Y = g_0(D, X) + \varepsilon, \quad E[\varepsilon | D, X] = 0
$$
$$
D = m_0(X) + V, \quad E[V | X] = 0
$$

where $m_0(X) = P(D=1|X)$ is the propensity score.

### Step 1: Generate Binary Treatment Data

```{r}
#| label: dgp-irm

# DGP with binary treatment
n <- 2000
p <- 10

X <- matrix(rnorm(n * p), n, p)
colnames(X) <- paste0("X", 1:p)

# Propensity score (log-odds model)
prop_score <- plogis(0.5 * X[,1] + 0.3 * X[,2] - 0.2 * X[,3])

# Binary treatment
D <- rbinom(n, 1, prop_score)

# Potential outcomes with heterogeneous effects
theta0_ate <- 1.0  # Average Treatment Effect
Y0 <- X[,1] + X[,2]^2 + rnorm(n)
Y1 <- Y0 + theta0_ate + 0.5 * X[,1]  # HTE component
Y <- D * Y1 + (1 - D) * Y0

irm_data <- data.table(Y = Y, D = D, X)
cat("True ATE =", theta0_ate, "\n")
cat("Treatment prevalence:", mean(D), "\n")
```

### Step 2: Create DoubleMLData for Binary Treatment

```{r}
#| label: irm-data

dml_irm_data <- DoubleMLData$new(
  data = irm_data,
  y_col = "Y",
  d_cols = "D",
  x_cols = paste0("X", 1:p)
)
```

### Step 3: Fit IRM with ATE Score

```{r}
#| label: irm-fit

# Outcome model: regression learner
ml_g <- lrn("regr.ranger", num.trees = 500, min.node.size = 5)

# Propensity score model: CLASSIFICATION learner with probability predictions
ml_m <- lrn("classif.ranger", num.trees = 500, min.node.size = 5,
            predict_type = "prob")

# Fit IRM for ATE
dml_irm <- DoubleMLIRM$new(
  data = dml_irm_data,
  ml_g = ml_g,
  ml_m = ml_m,
  n_folds = 5,
  score = "ATE"
)

dml_irm$fit()
dml_irm$summary()
```

::: callout-warning
## ⚠️ Common Pitfall: `predict_type = "prob"`

The propensity score model **must** use `predict_type = "prob"` to return probabilities, not class labels. Forgetting this is the #1 error students make with `DoubleMLIRM`. If you see an error about predictions not being in [0,1], check this first.
:::

### Step 4: Check Propensity Score Overlap

```{r}
#| label: overlap-check

# Extract propensity scores from the fitted model
# We'll compute them manually for the histogram
ps_model <- ranger::ranger(
  x = X, y = factor(D),
  num.trees = 500, probability = TRUE
)
ps_hat <- ps_model$predictions[, "1"]

# Overlap histogram
ps_df <- data.frame(ps = ps_hat, Treatment = factor(D, labels = c("Control", "Treated")))

ggplot(ps_df, aes(x = ps, fill = Treatment)) +
  geom_histogram(alpha = 0.5, bins = 40, position = "identity") +
  labs(
    title = "Propensity Score Distribution by Treatment Group",
    subtitle = "Good overlap = distributions overlap substantially",
    x = "Estimated Propensity Score",
    y = "Count"
  ) +
  theme_minimal(base_size = 14) +
  scale_fill_manual(values = c("steelblue", "coral"))
```

::: callout-note
## 🤔 Think About It

Look at the overlap histogram. Are there regions where one group dominates? Propensity scores near 0 or 1 create **positivity violations** — observations with extreme weights that inflate variance. DML mitigates this somewhat through cross-fitting, but severe overlap failures still cause problems.
:::

### Step 5: Compare ATE vs ATTE

```{r}
#| label: ate-vs-atte

# Fit with ATTE score
dml_irm_atte <- DoubleMLIRM$new(
  data = dml_irm_data,
  ml_g = lrn("regr.ranger", num.trees = 500, min.node.size = 5),
  ml_m = lrn("classif.ranger", num.trees = 500, min.node.size = 5,
             predict_type = "prob"),
  n_folds = 5,
  score = "ATTE"
)

dml_irm_atte$fit()

cat("ATE estimate: ", dml_irm$coef, " (SE:", dml_irm$se, ")\n")
cat("ATTE estimate:", dml_irm_atte$coef, " (SE:", dml_irm_atte$se, ")\n")
cat("True ATE:     ", theta0_ate, "\n")
```

::: callout-note
## 🤔 Think About It

When do ATE and ATTE differ substantially? The ATTE asks: "What is the effect **among those who were actually treated**?" If treatment effects are heterogeneous (they are in our DGP — the effect depends on $X_1$), ATE and ATTE can diverge. Which is the right estimand depends on your policy question.
:::

------------------------------------------------------------------------

## Exercise 4: Real Data — 401(k) Participation

Now let's apply DML to a real dataset. The 401(k) example from Chernozhukov et al. examines how 401(k) eligibility affects net financial assets.

### Background

Workers don't choose 401(k) eligibility directly — it depends on employer decisions. But eligibility is correlated with income, age, education, and other factors that also affect savings. DML lets us control for these flexibly.

### Step 1: Load the Data

```{r}
#| label: load-401k

# DoubleML provides the 401(k) dataset
data_401k <- fetch_401k()

# What do we have?
cat("Observations:", nrow(data_401k$data), "\n")
cat("Variables:", ncol(data_401k$data), "\n")

# Key variables:
# net_tfa   - Net total financial assets (outcome Y)
# e401      - 401(k) eligibility (treatment D)
# age, inc, educ, fsize, marr, ... (covariates X)

print(data_401k)
```

### Step 2: PLR — Effect of Eligibility on Assets

```{r}
#| label: 401k-plr

# Fit PLR with random forest
ml_l_401k <- lrn("regr.ranger", num.trees = 500, min.node.size = 5)
ml_m_401k <- lrn("regr.ranger", num.trees = 500, min.node.size = 5)

plr_401k <- DoubleMLPLR$new(
  data = data_401k,
  ml_l = ml_l_401k,
  ml_m = ml_m_401k,
  n_folds = 5,
  n_rep = 1
)

plr_401k$fit()
plr_401k$summary()
```

### Step 3: IRM — Binary Treatment Version

```{r}
#| label: 401k-irm

# Since eligibility is binary, IRM is more appropriate
ml_g_401k <- lrn("regr.ranger", num.trees = 500, min.node.size = 5)
ml_m_401k <- lrn("classif.ranger", num.trees = 500, min.node.size = 5,
                  predict_type = "prob")

irm_401k <- DoubleMLIRM$new(
  data = data_401k,
  ml_g = ml_g_401k,
  ml_m = ml_m_401k,
  n_folds = 5,
  score = "ATE"
)

irm_401k$fit()
irm_401k$summary()

cat("\nPLR estimate: ", plr_401k$coef, " (SE:", plr_401k$se, ")\n")
cat("IRM estimate: ", irm_401k$coef, " (SE:", irm_401k$se, ")\n")
```

::: callout-note
## 🤔 Think About It

The published estimates (Chernozhukov et al. 2018) find that 401(k) eligibility increases net financial assets by roughly \$8,000–\$10,000. How do your estimates compare? Why might PLR and IRM give different results for a binary treatment?

**Key point:** For binary treatments, IRM is generally preferred because PLR assumes a constant additive effect, while IRM allows the outcome model to differ between treated and control groups.
:::

### Step 4: Sensitivity to Learner and Folds

```{r}
#| label: 401k-sensitivity

# Try with different number of folds
for (k in c(3, 5, 10)) {
  irm_k <- DoubleMLIRM$new(
    data = data_401k,
    ml_g = lrn("regr.ranger", num.trees = 500, min.node.size = 5),
    ml_m = lrn("classif.ranger", num.trees = 500, min.node.size = 5,
               predict_type = "prob"),
    n_folds = k,
    score = "ATE"
  )
  irm_k$fit()
  cat("K =", k, "folds | Estimate:", round(irm_k$coef, 1),
      "| SE:", round(irm_k$se, 1), "\n")
}
```

### Step 5: Multiple Repetitions for Stability

```{r}
#| label: 401k-repetitions

# Use multiple repetitions of the cross-fitting procedure
# This reduces sensitivity to a particular random fold split
irm_rep <- DoubleMLIRM$new(
  data = data_401k,
  ml_g = lrn("regr.ranger", num.trees = 500, min.node.size = 5),
  ml_m = lrn("classif.ranger", num.trees = 500, min.node.size = 5,
             predict_type = "prob"),
  n_folds = 5,
  n_rep = 10,
  score = "ATE"
)

irm_rep$fit()
irm_rep$summary()

cat("\nWith 10 repetitions, the estimate is more stable.\n")
cat("Single rep:  ", round(irm_401k$coef, 1), "\n")
cat("10 reps:     ", round(irm_rep$coef, 1), "\n")
```

::: callout-tip
## Best Practice: Use Multiple Repetitions

In practice, always use `n_rep > 1` (typically 5–10). Each repetition uses a different random partition into folds. The final estimate averages across all repetitions, reducing sensitivity to the specific fold assignment. This costs more compute time but gives more reliable results.
:::

------------------------------------------------------------------------

## Wrap-Up & Connection to PS-2

### Key Takeaways

1. **DoubleML handles the plumbing** — cross-fitting, orthogonal scores, and valid inference are automated. You focus on choosing learners and specifying the model.

2. **Learner choice matters less than you'd think** — thanks to orthogonality, moderate nuisance estimation errors don't contaminate $\hat{\theta}$. But learners must be "good enough" (converge at $n^{-1/4}$ rate).

3. **PLR vs. IRM** — use PLR for continuous/discrete treatments with additive effects; use IRM for binary treatments where treatment-control outcome functions may differ.

4. **Always check overlap** — propensity scores near 0 or 1 signal positivity violations. No statistical method can fix a lack of overlap.

5. **Use multiple repetitions** — `n_rep = 5` or `n_rep = 10` stabilizes results against fold-assignment randomness.

### Common Debugging Tips

| Problem | Likely Cause | Fix |
|---------|-------------|-----|
| Error about predictions not in [0,1] | Missing `predict_type = "prob"` on classification learner | Add `predict_type = "prob"` to `classif.*` learner |
| Very large SEs | Poor overlap or weak first-stage predictions | Check propensity score histogram; try different learner |
| Estimate far from expected | Misspecified `y_col`, `d_cols`, or `x_cols` | Verify your `DoubleMLData` object with `print()` |
| "Column not found" error | Column names in data don't match specification | Check `colnames()` of your data |
| Results change dramatically across runs | Random fold assignment sensitivity | Increase `n_rep` (e.g., 10) |

### Connection to PS-2

Problem Set 2 asks you to apply DML to a **Medicaid expansion** scenario. The core workflow is identical to what we did today:

1. Create `DoubleMLData` with appropriate variable roles
2. Choose and configure ML learners
3. Fit both PLR and IRM models
4. Compare learners and conduct sensitivity analysis
5. Interpret results in policy context

The key difference: PS-2 uses a more realistic dataset where you'll need to think carefully about which covariates to include and which estimand (ATE vs ATTE) is policy-relevant.

### Looking Ahead

**Session 2.5 (Wed Mar 11):** Doubly-Robust Estimation — AIPW & TMLE

- How DML relates to the broader doubly-robust estimation literature
- When you want AIPW vs. DML
- The `tmle` and `drtmle` packages

------------------------------------------------------------------------

## Additional Resources

- [DoubleML R Documentation](https://docs.doubleml.org/r/stable/)
- [DoubleML 401(k) Vignette](https://docs.doubleml.org/stable/examples/R_double_ml_bonus.html)
- Bach et al. (2021) — "DoubleML: An Object-Oriented Implementation of Double Machine Learning in R"
- [mlr3 Learner Reference](https://mlr3learners.mlr-org.com/)
