---
title: "PS 2: Code Companion"
subtitle: "Reusable Patterns for Double Machine Learning"
author: "HPM 883: Advanced Quantitative Methods"
date: "Spring 2026"
format:
  html:
    toc: true
    toc-depth: 2
    code-tools: true
    code-fold: false
execute:
  eval: false
  echo: true
  warning: false
editor: visual
---

::: {.callout-important}
## How to Use This Document

This companion demonstrates DML code patterns using **practice data** — not the
PS2 Medicaid dataset. You'll need to **adapt** variable names, column selections,
and learner choices to fit your problem set.

**Workflow:**

1. Read the relevant section for the analysis you're attempting
2. Run the example code to see expected output shapes
3. Adapt the pattern for your PS2 data and questions
:::

## Package Installation

If you haven't installed the DML packages yet, run this once:

```{r}
#| label: install-packages
#| eval: false

install.packages(c(
  "DoubleML", "mlr3", "mlr3learners",
  "ranger", "xgboost", "glmnet",
  "cobalt", "data.table", "ggplot2"
))
```

## Setup: Practice Data

We use DoubleML's built-in data generator to create practice data with a known
treatment effect. This has **different variables** than your PS2 data — the point
is to learn the patterns, not to copy the code verbatim.

```{r}
#| label: setup

library(DoubleML)
library(mlr3)
library(mlr3learners)
library(data.table)
library(ggplot2)

# Generate practice data: 2000 obs, 20 covariates
# True parameter: theta_0 = 0.5 (known)
set.seed(42)
practice_dt <- make_plr_CCDDHNR2018(
  n_obs = 2000,
  dim_x = 20,
  return_type = "data.table"
)

# Peek at the structure
str(practice_dt)
```

::: {.callout-note}
The practice data has columns `y` (outcome), `d` (treatment), and `X1`–`X20`
(confounders). Your PS2 data has `Y`, `D`, `Y_pre`, and `X1`–`X50`.
:::

---

## Getting Started with DoubleML

The `DoubleML` package requires a **data backend** object that specifies which
columns are the outcome, treatment, and confounders.

```{r}
#| label: data-backend

# Create the DoubleML data backend
# Key arguments: y_col, d_cols, x_cols
dml_data <- DoubleMLData$new(
  data = practice_dt,
  y_col = "y",
  d_cols = "d",
  x_cols = paste0("X", 1:20)
)

# Check the setup
print(dml_data)
```

::: {.callout-warning}
## Gotcha: Column Types

`DoubleMLData$new()` expects a `data.table` or `data.frame`. If your data is a
tibble, convert it first: `as.data.table(your_data)`.

Also make sure `d_cols` contains the **treatment** column name (not the outcome).
A common mistake is swapping `y_col` and `d_cols`.
:::

---

## Fitting a PLR Model

The **Partially Linear Regression** model assumes:

$$Y = D\tau + g(X) + U, \quad D = m(X) + V$$

You need two ML learners: one for the outcome (`ml_l`) and one for the treatment
(`ml_m`). For PLR, **both are regression learners** (even when treatment is binary).

```{r}
#| label: plr-basic

# Step 1: Create learners
ml_l <- lrn("regr.cv_glmnet")   # outcome model (Lasso)
ml_m <- lrn("regr.cv_glmnet")   # treatment model (Lasso)

# Step 2: Create PLR object
set.seed(42)
dml_plr <- DoubleMLPLR$new(
  data = dml_data,
  ml_l = ml_l,
  ml_m = ml_m,
  n_folds = 5,
  score = "partialling out"
)

# Step 3: Fit
dml_plr$fit()

# Step 4: Results
dml_plr$summary()
```

**Extracting specific values:**

```{r}
#| label: plr-extract

# Point estimate
cat("ATE:", as.numeric(dml_plr$coef), "\n")

# Standard error
cat("SE:", as.numeric(dml_plr$se), "\n")

# Confidence interval (returns a matrix)
ci <- dml_plr$confint()
cat("95% CI: [", ci[1, 1], ",", ci[1, 2], "]\n")
```

---

## Fitting an IRM Model

The **Interactive Regression Model** allows treatment effect heterogeneity:

$$Y = g(D, X) + U, \quad D = m(X) + V$$

For IRM, the treatment model must be a **classification** learner since $D$ is
binary. The outcome model stays as regression.

```{r}
#| label: irm-basic

# For IRM, we need binary treatment. Binarize the practice data.
practice_dt_irm <- copy(practice_dt)
practice_dt_irm[, d := as.integer(d > median(d))]

dml_data_irm <- DoubleMLData$new(
  data = practice_dt_irm,
  y_col = "y",
  d_cols = "d",
  x_cols = paste0("X", 1:20)
)

# CRITICAL: Use classif. learner with predict_type = "prob"
ml_g <- lrn("regr.cv_glmnet")                              # outcome
ml_m_irm <- lrn("classif.cv_glmnet", predict_type = "prob") # treatment

set.seed(42)
dml_irm <- DoubleMLIRM$new(
  data = dml_data_irm,
  ml_g = ml_g,
  ml_m = ml_m_irm,
  n_folds = 5,
  score = "ATE"
)

dml_irm$fit()
dml_irm$summary()
```

::: {.callout-warning}
## Gotcha: `classif.` Prefix

For the IRM treatment model, you **must** use `classif.` learners (not `regr.`):

- `lrn("classif.cv_glmnet", predict_type = "prob")` — Lasso classification
- `lrn("classif.ranger", predict_type = "prob")` — Random Forest classification
- `lrn("classif.xgboost", predict_type = "prob")` — XGBoost classification

And you **must** set `predict_type = "prob"` so the model returns probabilities
(propensity scores), not just class labels.

If you forget `predict_type = "prob"`, you'll get an error about predictions
not being in `[0, 1]`.
:::

---

## Comparing ML Learners

To compare different ML methods, fit the same model structure with different
learner pairs and collect the results.

```{r}
#| label: learner-comparison

# Define three learner pairs
learner_pairs <- list(
  Lasso = list(
    ml_l = lrn("regr.cv_glmnet"),
    ml_m = lrn("regr.cv_glmnet")
  ),
  `Random Forest` = list(
    ml_l = lrn("regr.ranger", num.trees = 500),
    ml_m = lrn("regr.ranger", num.trees = 500)
  ),
  XGBoost = list(
    ml_l = lrn("regr.xgboost", nrounds = 100, verbose = 0),
    ml_m = lrn("regr.xgboost", nrounds = 100, verbose = 0)
  )
)

# Fit each and collect results
results <- lapply(names(learner_pairs), function(name) {
  set.seed(42)
  obj <- DoubleMLPLR$new(
    data = dml_data,
    ml_l = learner_pairs[[name]]$ml_l,
    ml_m = learner_pairs[[name]]$ml_m,
    n_folds = 5
  )
  obj$fit()
  ci <- obj$confint()
  data.frame(
    Learner = name,
    Estimate = as.numeric(obj$coef),
    SE = as.numeric(obj$se),
    CI_Lower = ci[1, 1],
    CI_Upper = ci[1, 2]
  )
})

comparison <- do.call(rbind, results)
knitr::kable(comparison, digits = 4)
```

---

## Creating a Forest Plot

A forest plot visually compares estimates and confidence intervals across methods.

```{r}
#| label: forest-plot
#| fig-width: 8
#| fig-height: 4

ggplot(comparison, aes(x = Estimate, y = Learner)) +
  geom_pointrange(aes(xmin = CI_Lower, xmax = CI_Upper), size = 0.8) +
  geom_vline(
    xintercept = 0.5,  # true value for practice data
    linetype = "dashed", color = "red", alpha = 0.7
  ) +
  labs(
    x = "Treatment Effect Estimate",
    y = "",
    title = "Learner Comparison: Forest Plot",
    caption = "Dashed line = true parameter value"
  ) +
  theme_minimal(base_size = 13)
```

---

## Extracting Propensity Scores

To diagnose overlap, you need individual-level propensity scores. The cleanest
approach is to fit a separate cross-validated propensity model.

```{r}
#| label: propensity-scores

library(ranger)

# Cross-fitted propensity scores
n <- nrow(practice_dt_irm)
folds <- sample(rep(1:5, length.out = n))
e_hat <- numeric(n)

X_cols <- paste0("X", 1:20)
X_mat <- practice_dt_irm[, ..X_cols]

for (k in 1:5) {
  train_idx <- which(folds != k)
  test_idx <- which(folds == k)

  train_df <- data.frame(
    D = factor(practice_dt_irm$d[train_idx]),
    X_mat[train_idx, ]
  )
  test_df <- data.frame(X_mat[test_idx, ])

  fit <- ranger(D ~ ., data = train_df,
    probability = TRUE, num.trees = 500)
  e_hat[test_idx] <- predict(fit, data = test_df)$predictions[, "1"]
}
```

```{r}
#| label: propensity-histogram
#| fig-width: 8
#| fig-height: 5

ps_df <- data.frame(
  e_hat = e_hat,
  D = factor(practice_dt_irm$d, labels = c("Control", "Treated"))
)

ggplot(ps_df, aes(x = e_hat, fill = D)) +
  geom_histogram(alpha = 0.5, position = "identity", bins = 40) +
  scale_fill_manual(values = c("Control" = "#3498DB", "Treated" = "#E74C3C")) +
  labs(
    x = "Estimated Propensity Score",
    y = "Count",
    title = "Propensity Score Distribution by Treatment Status",
    fill = "Group"
  ) +
  theme_minimal(base_size = 13)
```

---

## Propensity Score Trimming

Drop observations with extreme propensity scores and re-fit.

```{r}
#| label: trimming

# Define trimming thresholds
trim_lo <- 0.05
trim_hi <- 0.95

keep <- e_hat >= trim_lo & e_hat <= trim_hi
cat(sprintf("Dropped: %d obs (%.1f%%)\n", sum(!keep), 100 * mean(!keep)))

# Re-fit DML on trimmed sample
trimmed_dt <- practice_dt_irm[keep, ]
dml_trimmed <- DoubleMLData$new(
  data = trimmed_dt,
  y_col = "y", d_cols = "d",
  x_cols = paste0("X", 1:20)
)

set.seed(42)
plr_trimmed <- DoubleMLPLR$new(
  dml_trimmed,
  lrn("regr.cv_glmnet"), lrn("regr.cv_glmnet"),
  n_folds = 5
)
plr_trimmed$fit()

cat("Full sample estimate:", as.numeric(dml_plr$coef), "\n")
cat("Trimmed estimate:", as.numeric(plr_trimmed$coef), "\n")
```

---

## Overlap Weighting

An alternative to trimming: weight each observation by $w_i = \hat{e}(X_i)(1 - \hat{e}(X_i))$. This smoothly downweights extreme propensity scores.

```{r}
#| label: overlap-weights

# Overlap weights
w <- e_hat * (1 - e_hat)

# Weighted difference in means (Hajek estimator)
d_vec <- practice_dt_irm$d
y_vec <- practice_dt_irm$y

ate_overlap <-
  sum(w * d_vec * y_vec) / sum(w * d_vec) -
  sum(w * (1 - d_vec) * y_vec) / sum(w * (1 - d_vec))

cat("Overlap-weighted ATE:", round(ate_overlap, 4), "\n")
```

---

## Covariate Balance with cobalt

The `cobalt` package provides easy balance diagnostics.

```{r}
#| label: cobalt-balance

library(cobalt)

# IPW weights for balance check
ipw <- ifelse(d_vec == 1, 1 / e_hat, 1 / (1 - e_hat))

# Focus on first 10 covariates for readability
bal_vars <- practice_dt_irm[, paste0("X", 1:10), with = FALSE]
```

```{r}
#| label: love-plot
#| fig-width: 8
#| fig-height: 5

love.plot(
  bal_vars,
  treat = practice_dt_irm$d,
  weights = ipw,
  method = "weighting",
  un = TRUE,
  threshold = 0.1,
  abs = TRUE,
  var.order = "unadjusted",
  title = "Covariate Balance: Unadjusted vs IPW"
) +
  theme_minimal()
```

```{r}
#| label: balance-table

bal.tab(
  bal_vars,
  treat = practice_dt_irm$d,
  weights = ipw,
  method = "weighting",
  un = TRUE
)
```

---

## Manual AIPW Estimator

The AIPW estimator combines outcome modeling and propensity weighting. You need
three predictions per observation (all cross-fitted):

1. $\hat{\mu}_1(X)$ — predicted outcome under treatment
2. $\hat{\mu}_0(X)$ — predicted outcome under control
3. $\hat{e}(X)$ — predicted propensity score

```{r}
#| label: manual-aipw

set.seed(42)
mu1 <- numeric(n)
mu0 <- numeric(n)
e_aipw <- numeric(n)

for (k in 1:5) {
  train_idx <- which(folds != k)
  test_idx <- which(folds == k)
  d_train <- practice_dt_irm$d[train_idx]
  y_train <- practice_dt_irm$y[train_idx]

  # Outcome under treatment (fit on D=1 only)
  idx_1 <- train_idx[d_train == 1]
  fit_1 <- ranger(y ~ .,
    data = data.frame(y = practice_dt_irm$y[idx_1], X_mat[idx_1, ]),
    num.trees = 500)
  mu1[test_idx] <- predict(fit_1, data = data.frame(X_mat[test_idx, ]))$predictions

  # Outcome under control (fit on D=0 only)
  idx_0 <- train_idx[d_train == 0]
  fit_0 <- ranger(y ~ .,
    data = data.frame(y = practice_dt_irm$y[idx_0], X_mat[idx_0, ]),
    num.trees = 500)
  mu0[test_idx] <- predict(fit_0, data = data.frame(X_mat[test_idx, ]))$predictions

  # Propensity model
  fit_e <- ranger(D ~ .,
    data = data.frame(D = factor(d_train), X_mat[train_idx, ]),
    probability = TRUE, num.trees = 500)
  e_aipw[test_idx] <- predict(fit_e, data = data.frame(X_mat[test_idx, ]))$predictions[, "1"]
}

# Clip extreme propensity scores
e_aipw <- pmax(0.01, pmin(0.99, e_aipw))

# AIPW formula
y_vec <- practice_dt_irm$y
d_vec <- practice_dt_irm$d

psi <- mu1 - mu0 +
  d_vec * (y_vec - mu1) / e_aipw -
  (1 - d_vec) * (y_vec - mu0) / (1 - e_aipw)

tau_aipw <- mean(psi)
se_aipw <- sd(psi) / sqrt(n)
ci_lo <- tau_aipw - 1.96 * se_aipw
ci_hi <- tau_aipw + 1.96 * se_aipw

cat(sprintf("AIPW Estimate: %.4f (SE: %.4f)\n", tau_aipw, se_aipw))
cat(sprintf("95%% CI: [%.4f, %.4f]\n", ci_lo, ci_hi))
```

::: {.callout-note}
## Key Steps in the AIPW Loop

1. **Split** data into K folds
2. **For each fold:** fit three models on the *other* folds, predict on the held-out fold
3. **Combine** all predictions into the AIPW influence function $\hat{\psi}_i$
4. **Estimate:** $\hat{\tau} = \bar{\psi}$, $\hat{SE} = sd(\hat{\psi}) / \sqrt{n}$

The cross-fitting is essential — using full-sample predictions would introduce
overfitting bias.
:::

---

## Code Skeletons

::: {.callout-warning}
## Try First, Then Use These

These code skeletons provide more detailed scaffolding for specific PS-2 tasks.
**Only use them after you've attempted the task on your own** — try adapting the
general patterns above first. If you're still stuck after 30+ minutes, these
skeletons can help you move forward.
:::

### Task 2.1: PLR with Lasso

```{r}
#| label: skeleton-2-1
#| eval: false

dml_data <- DoubleMLData$new(
  data = as.data.table(___),
  y_col = "___",
  d_cols = "___",
  x_cols = c(paste0("X", 1:50), "Y_pre")
)

ml_l <- lrn("regr.cv_glmnet")
ml_m <- lrn("regr.cv_glmnet")

set.seed(883)
dml_plr <- DoubleMLPLR$new(dml_data, ml_l, ml_m, n_folds = ___)
dml_plr$___()
dml_plr$summary()
```

### Task 2.2: Learner Comparison

```{r}
#| label: skeleton-2-2
#| eval: false

learner_pairs <- list(
  Lasso = list(ml_l = lrn("regr.cv_glmnet"), ml_m = lrn("regr.cv_glmnet")),
  RF    = list(ml_l = lrn("regr.ranger", num.trees = 500),
               ml_m = lrn("regr.ranger", num.trees = 500)),
  XGB   = list(ml_l = lrn("regr.xgboost", nrounds = 100, verbose = 0),
               ml_m = lrn("regr.xgboost", nrounds = 100, verbose = 0))
)

results <- lapply(names(learner_pairs), function(name) {
  set.seed(883)
  obj <- DoubleMLPLR$new(dml_data, ___, ___, n_folds = 5)
  obj$fit()
  ci <- obj$confint()
  data.frame(Learner = name, Estimate = as.numeric(obj$coef),
             CI_Lower = ci[1,1], CI_Upper = ci[1,2])
})

comparison <- do.call(rbind, results)
# Then use ggplot + geom_pointrange for the forest plot
```

### Task 2.3: IRM

```{r}
#| label: skeleton-2-3
#| eval: false

ml_g <- lrn("regr.cv_glmnet")          # outcome model (regression)
ml_m <- lrn("classif.cv_glmnet",        # treatment model (classification!)
             predict_type = "prob")      # must return probabilities

set.seed(883)
dml_irm <- DoubleMLIRM$new(dml_data, ml_g, ml_m, n_folds = 5, score = "ATE")
dml_irm$fit()
dml_irm$summary()
```

### Task 4.1: Manual AIPW

```{r}
#| label: skeleton-4-1
#| eval: false

K <- 5
folds <- sample(rep(1:K, length.out = nrow(medicaid_data)))
mu1_hat <- mu0_hat <- e_hat <- numeric(nrow(medicaid_data))

for (k in 1:K) {
  train <- which(folds != k)
  test  <- which(folds == k)

  # Fit outcome model on treated obs in training fold
  # Fit outcome model on control obs in training fold
  # Fit propensity model on all training obs
  # Predict on test fold
}

# Clip e_hat to avoid division by zero
e_hat <- pmax(0.01, pmin(0.99, e_hat))

# AIPW influence function
psi <- mu1_hat - mu0_hat +
  D * (Y - mu1_hat) / e_hat -
  (1 - D) * (Y - mu0_hat) / (1 - e_hat)

tau_aipw <- mean(psi)
se_aipw  <- sd(psi) / sqrt(length(psi))
```

---

## Troubleshooting

::: {.callout-warning}
## Common Errors and Fixes

**"Column 'd' not found"** or similar:

- Check that `y_col`, `d_cols`, `x_cols` match your actual column names exactly
- R is case-sensitive: `"Y"` ≠ `"y"`, `"D"` ≠ `"d"`

**"Column of class 'character' is not supported":**

- Convert character columns to numeric or factor before creating the data backend

**"Error in lrn(...)": Learner not found:**

- Make sure `mlr3learners` is loaded: `library(mlr3learners)`
- For Lasso: `"regr.cv_glmnet"` (not `"regr.glmnet"`)
- For Random Forest: `"regr.ranger"` (not `"regr.randomForest"`)
- For XGBoost: `"regr.xgboost"` (needs `xgboost` package installed)

**IRM errors about predictions not in [0,1]:**

- You need `classif.` learners (not `regr.`) for the treatment model
- Must include `predict_type = "prob"`

**"Cannot coerce to data.table":**

- Wrap your data: `as.data.table(your_dataframe)`

**Very slow fitting:**

- Reduce `num.trees` for ranger (try 200 instead of 500)
- Reduce `nrounds` for xgboost (try 50)
- Reduce `n_folds` from 5 to 3 (trades some bias for speed)
:::
