---
title: "Lab 4: Causal Forests in Practice"
subtitle: "Code-Along Lab"
author: "Sean Sylvia"
date: "April 1, 2026"
slug: "lab-4-causal-forests"
categories: [Lab, Code-Along, Causal Forests, grf, HTE]
description: "Hands-on implementation of the causal forest analysis pipeline using grf. Estimate CATEs, test for real heterogeneity, summarize with GATES/CLAN, evaluate targeting with RATE, and preview policy trees."
format:
  html:
    toc: true
    toc-depth: 2
    code-tools: true
editor: visual
execute:
  eval: true
---

## Accept the Assignment

[**Click here to accept Lab 4 on GitHub Classroom**](https://classroom.github.com/a/CDcZhKw9)

This will create your personal repository for this lab.

**To get started:**

1.  Accept the assignment using the link above
2.  Clone your repository locally: `git clone <your-repo-url>`
3.  Open `hpm883-lab.Rproj` in RStudio or Positron
4.  Run `renv::restore()` to install packages
5.  Install the causal forest packages (not yet in renv): `install.packages(c("grf", "policytree", "DiagrammeR"))`
6.  Open `lab-4-causal-forests.qmd` and follow along

::: callout-important
## Code-Along Session (Remote, Zoom)

This is a **75-minute guided session**. Code chunks execute when rendered, but we recommend running them interactively in RStudio or Positron to explore the output step by step.
:::

## Overview

In Session 3.0, we introduced causal forests and the GRF analysis pipeline:

```         
1. Estimate CATEs with causal_forest()
2. Test whether heterogeneity is real (test_calibration)
3. Screen which covariates drive heterogeneity (variable_importance)
4. Summarize patterns (best_linear_projection)
5. Visualize who benefits most (GATES)
6. Characterize high-benefit groups (CLAN)
7. Evaluate targeting quality (RATE/AUTOC)
8. Preview policy rules (policy_tree)
```

Today we implement this pipeline hands-on, including a **failure case** where we deliberately simulate data with no heterogeneity to see how the diagnostics catch it. By the end of this session, you will:

1.  Fit a causal forest and extract out-of-bag CATE predictions
2.  Use `test_calibration()` to distinguish real heterogeneity from noise
3.  Build GATES and CLAN analyses from AIPW scores
4.  Evaluate targeting quality with RATE/AUTOC
5.  Preview `policytree` for interpretable treatment rules

------------------------------------------------------------------------

## Setup & Orientation

### Load Packages

```{r}
#| label: setup
#| message: false
#| warning: false

# Core causal forest packages
library(grf)
library(policytree)

# Visualization and inference
library(ggplot2)
library(sandwich)
library(lmtest)

# Reproducibility
set.seed(883)

# Plot theme
theme_set(theme_minimal(base_size = 14))
```

### The GRF Architecture

`grf` (Generalized Random Forests) estimates heterogeneous treatment effects using a forest-based approach. The key insight: each tree splits the data to maximize treatment effect *heterogeneity* across leaves, not prediction accuracy.

| Component | Function | Purpose |
|--------------------------|------------------------|----------------------|
| **Estimation** | `causal_forest()` | Fit forest, estimate individual CATEs |
| **Diagnostics** | `test_calibration()` | Test if heterogeneity is real |
| **Screening** | `variable_importance()` | Which covariates drive heterogeneity |
| **Summary** | `best_linear_projection()` | Interpretable CATE coefficients |
| **Evaluation** | `rank_average_treatment_effect()` | Targeting quality (AUTOC) |
| **Policy** | `policy_tree()` | Simple, interpretable treatment rules |

The workflow is always: **Estimate → Test → Summarize → Evaluate → Decide**.

::: callout-tip
## How `grf` differs from regular random forests

Standard random forests optimize *prediction accuracy* (minimize MSE). Causal forests optimize *treatment effect heterogeneity* — they split on covariates that create the largest *difference in treatment effects* across children. This is why variable importance in `grf` tells you about effect *modifiers*, not outcome predictors.
:::

### The Narrative: CHW Program

**Setting:** A community health worker (CHW) program aims to reduce emergency room visits among patients with chronic conditions. A randomized trial assigned 2,000 patients to receive CHW support or usual care.

The trial found an overall reduction in ER visits, but the program is expensive. Hospital leadership wants to know: **Who benefits most?** Can we target the program to maximize impact within a limited budget?

We'll use simulated data where we *know* the true treatment effect function, so we can verify our estimates.

------------------------------------------------------------------------

## Exercise 1: Estimate CATEs — The Success Case

### Step 1: Simulate the CHW Trial Data

We create data where treatment effects genuinely vary: patients with more comorbidities and older age benefit more from CHW support.

```{r}
#| label: simulate-data

n <- 2000
p <- 6

# Patient characteristics
age <- runif(n, 25, 75)
comorbidities <- rpois(n, lambda = 1.5)
female <- rbinom(n, 1, 0.52)
income_quartile <- sample(1:4, n, replace = TRUE)
rural <- rbinom(n, 1, 0.35)
prior_er_visits <- rpois(n, lambda = 3)

X <- data.frame(
  age = age,
  comorbidities = comorbidities,
  female = female,
  income_quartile = income_quartile,
  rural = rural,
  prior_er_visits = prior_er_visits
)

# Treatment assignment (RCT: 50/50)
W <- rbinom(n, 1, 0.5)

# True CATE function: effect depends on comorbidities and age
# Higher comorbidities -> larger benefit (more ER visits prevented)
# Older patients -> larger benefit
# Base effect is small
tau.true <- -0.5 + -0.8 * (comorbidities >= 2) + -0.6 * (age > 55)

# Outcome: ER visits in 12 months
Y <- 4 + 0.3 * comorbidities + 0.02 * age - 0.5 * female +
  tau.true * W + rnorm(n, 0, 2)

cat("True ATE:", round(mean(tau.true), 3), "\n")
cat("True CATE range:", round(range(tau.true), 3), "\n")
cat("Fraction with large benefit (tau < -1):", mean(tau.true < -1), "\n")
```

::: callout-note
## Understanding the DGP

The true treatment effect has **four possible values** depending on comorbidities and age:

| Comorbidities |    Age    | True $\tau(x)$ |
|:-------------:|:---------:|:--------------:|
|     \< 2      | $\leq$ 55 |      -0.5      |
|   $\geq$ 2    | $\leq$ 55 |      -1.3      |
|     \< 2      |   \> 55   |      -1.1      |
|   $\geq$ 2    |   \> 55   |      -1.9      |

Everyone benefits (all effects are negative = fewer ER visits), but the benefit ranges from modest (-0.5) to substantial (-1.9). The four other covariates (`female`, `income_quartile`, `rural`, `prior_er_visits`) are **not** effect modifiers — they don't appear in the CATE function.
:::

### Step 2: Fit the Causal Forest

```{r}
#| label: fit-causal-forest

# Fit a causal forest
# W.hat = 0.5 because this is an RCT with known assignment probability
cf <- causal_forest(
  X = as.matrix(X),
  Y = Y,
  W = W,
  W.hat = 0.5,        # Known propensity (RCT)
  num.trees = 2000,
  seed = 883
)

# Out-of-bag CATE predictions
tau.hat <- predict(cf)$predictions

# Forest-weighted ATE
cat("Estimated ATE (forest-weighted):\n")
average_treatment_effect(cf)
```

::: callout-note
## Why `W.hat = 0.5`?

In an RCT, we *know* the propensity score: every patient had a 50% chance of treatment. Setting `W.hat = 0.5` tells the forest to skip propensity estimation, which both speeds up computation and avoids unnecessary noise. In observational studies, you'd omit this and let the forest estimate propensities.
:::

### Step 3: Visualize the CATE Distribution

```{r}
#| label: cate-histogram
#| fig-alt: "Histogram of estimated CATEs showing distribution of predicted treatment effects across individuals"

ggplot(data.frame(tau = tau.hat), aes(x = tau)) +
  geom_histogram(bins = 40, fill = "#4B9CD3", color = "white", alpha = 0.8) +
  geom_vline(xintercept = mean(tau.hat), color = "#E07C3E",
             linewidth = 1.2, linetype = "dashed") +
  labs(
    title = "Distribution of Estimated CATEs",
    subtitle = "Orange dashed line = ATE (forest-weighted average)",
    x = expression(hat(tau)(x)),
    y = "Count"
  ) +
  annotate("text", x = mean(tau.hat) + 0.15, y = Inf, vjust = 2,
           label = paste("ATE =", round(mean(tau.hat), 2)),
           color = "#E07C3E", fontface = "bold")
```

::: callout-important
## Think About It #1

Look at the histogram of estimated CATEs.

1.  The spread looks wider than you might expect from just four true values. Why? *(Hint: think about estimation noise in finite samples.)*
2.  How would you determine whether this spread reflects **real** heterogeneity vs. **estimation noise**?
3.  If someone showed you this histogram without any diagnostics, would you be confident the heterogeneity is real?

We'll answer question 2 formally in Exercise 2.
:::

::: callout-tip
## AI Guidance: CATE Histograms

If you ask an AI tool to "interpret this CATE histogram," it will almost certainly say "there is substantial heterogeneity." That's because the histogram always *looks* spread out — even when there's no real heterogeneity. **Never interpret a CATE histogram without running `test_calibration()` first.** The histogram shows estimated noise + signal; only diagnostics can separate the two.
:::

------------------------------------------------------------------------

## Exercise 2: Test & Summarize + The Failure Case

### Step 1: Calibration Test (Success Case)

The calibration test asks: "Does the forest's heterogeneity reflect real variation in treatment effects, or is it just noise?"

```{r}
#| label: test-calibration-success

cal <- test_calibration(cf)
print(cal)
```

::: callout-tip
## How to Read `test_calibration()`

The output has two rows:

1.  **mean.forest.prediction** — Tests whether the ATE estimate is calibrated. Coefficient $\approx 1$ means the forest's average prediction is on target.
2.  **differential.forest.prediction** — Tests whether heterogeneity is real. Coefficient \> 0 with p \< 0.05 means the forest detects **genuine** variation in treatment effects beyond noise.

The differential coefficient is the one that matters most. If it's not significant, your "heterogeneity" may be an artifact.
:::

### Step 2: Variable Importance

Which covariates does the forest rely on for splitting?

```{r}
#| label: variable-importance
#| fig-alt: "Bar chart showing variable importance scores for each covariate in the causal forest"

var.imp <- variable_importance(cf)
var.imp.df <- data.frame(
  variable = colnames(X),
  importance = as.numeric(var.imp)
)
var.imp.df <- var.imp.df[order(-var.imp.df$importance), ]

ggplot(var.imp.df, aes(x = reorder(variable, importance), y = importance)) +
  geom_col(fill = "#4B9CD3") +
  coord_flip() +
  labs(
    title = "Variable Importance for Treatment Effect Heterogeneity",
    subtitle = "How often each variable is used in forest splits",
    x = NULL, y = "Importance (weighted split frequency)"
  )
```

::: callout-warning
## Variable Importance $\neq$ Effect Direction

Variable importance tells you *which* covariates the forest splits on — not the direction or magnitude of the effect. `comorbidities` and `age` should dominate (they are the true effect modifiers). The other four covariates are noise — they should have low importance, but won't be exactly zero.

Use BLP (next step) for interpretable summaries of *how* covariates relate to CATEs.
:::

### Step 3: Best Linear Projection (BLP)

BLP projects the CATEs onto covariates of interest, giving interpretable coefficients with standard errors.

```{r}
#| label: blp

blp <- best_linear_projection(
  cf,
  A = as.matrix(X[, c("comorbidities", "age", "female")])
)
print(blp)
```

::: callout-note
## Interpreting BLP

Each coefficient tells you: "A one-unit increase in this covariate is associated with this much change in the treatment effect."

For example, if the comorbidities coefficient is -0.7, each additional comorbidity increases the treatment *benefit* by 0.7 fewer ER visits. The `female` coefficient should be near zero (it's not a true effect modifier in our DGP).

**This is NOT a causal claim about the covariate** — it's a summary of heterogeneity patterns. BLP answers "how do CATEs covary with X?" not "does X cause the effect to change?"
:::

### Step 4: The Failure Case — Constant Effects

Now let's see what happens when there's **no heterogeneity**. We simulate a constant treatment effect and show how the diagnostics catch it.

```{r}
#| label: constant-effect-dgp

# Same setup, but constant treatment effect for everyone
tau.constant <- rep(-1.0, n)  # Everyone gets the same effect

Y.const <- 4 + 0.3 * comorbidities + 0.02 * age - 0.5 * female +
  tau.constant * W + rnorm(n, 0, 2)

# Fit causal forest on constant-effect data
cf.const <- causal_forest(
  X = as.matrix(X),
  Y = Y.const,
  W = W,
  W.hat = 0.5,
  num.trees = 2000,
  seed = 883
)

tau.hat.const <- predict(cf.const)$predictions
```

```{r}
#| label: compare-histograms
#| fig-alt: "Side-by-side histograms comparing estimated CATE distributions under true heterogeneity versus constant effects"

# Compare the two histograms
par(mfrow = c(1, 2))

hist(tau.hat, breaks = 40, col = "#4B9CD3", border = "white",
     main = "Success: Real Heterogeneity",
     xlab = expression(hat(tau)(x)), xlim = c(-3, 1))
abline(v = mean(tau.hat), col = "#E07C3E", lwd = 2, lty = 2)

hist(tau.hat.const, breaks = 40, col = "#9B59B6", border = "white",
     main = "Failure: Constant Effect",
     xlab = expression(hat(tau)(x)), xlim = c(-3, 1))
abline(v = mean(tau.hat.const), col = "#E07C3E", lwd = 2, lty = 2)

par(mfrow = c(1, 1))
```

```{r}
#| label: test-calibration-failure

# Calibration test on the constant-effect forest
cal.const <- test_calibration(cf.const)
print(cal.const)

cat("\n--- Comparison ---\n")
cat("Success case - differential coef:",
    round(cal[2, 1], 3), "(p =", round(cal[2, 4], 4), ")\n")
cat("Failure case - differential coef:",
    round(cal.const[2, 1], 3), "(p =", round(cal.const[2, 4], 4), ")\n")
```

::: callout-important
## Think About It #2

Compare the two histograms and calibration test results.

1.  Both histograms show *spread* in estimated CATEs. Why does the constant-effect histogram still show variation? *(Hint: estimation noise.)*
2.  The calibration test catches the difference. What specifically changes between the success and failure case?
3.  If you only looked at histograms without running `test_calibration()`, would you have been fooled by the constant-effect case?

**Key lesson:** Visual spread in CATE estimates is *necessary but not sufficient* evidence of heterogeneity. Always run formal tests.
:::

------------------------------------------------------------------------

## Exercise 3: GATES + CLAN + RATE

### Step 1: GATES (Group Average Treatment Effects)

We rank individuals by predicted CATE, split into quintiles, and estimate the ATE within each group using AIPW scores.

```{r}
#| label: gates-setup

# Create quintile rankings using K-fold approach (avoids overfitting)
num.rankings <- 5
num.folds <- 5
folds <- sample(rep(1:num.folds, length.out = n))

ranking <- rep(NA, n)
for (fold in seq(num.folds)) {
  in.fold <- (folds == fold)
  tau.fold <- tau.hat[in.fold]
  breaks <- quantile(tau.fold, probs = seq(0, 1, by = 1 / num.rankings))
  ranking[in.fold] <- as.numeric(cut(tau.fold, breaks,
                                      include.lowest = TRUE,
                                      labels = seq(num.rankings)))
}

# Compute AIPW scores from forest components
e.hat <- cf$W.hat   # propensity (known = 0.5 in RCT)
m.hat <- cf$Y.hat   # outcome model E[Y|X]
mu.hat.0 <- m.hat - e.hat * tau.hat
mu.hat.1 <- m.hat + (1 - e.hat) * tau.hat

aipw.scores <- tau.hat +
  W / e.hat * (Y - mu.hat.1) -
  (1 - W) / (1 - e.hat) * (Y - mu.hat.0)

# OLS with quintile dummies (no intercept)
ols <- lm(aipw.scores ~ 0 + factor(ranking))
gates <- coeftest(ols, vcov = vcovHC(ols, "HC2"))
print(gates)
```

::: callout-note
## Why AIPW Scores?

We can't just compute the difference in means within each CATE quintile — that would be circular (the quintiles are defined by the same forest that generated the predictions). Instead, we use **AIPW scores**, which are doubly robust estimates of individual treatment effects. Regressing AIPW scores on quintile dummies gives us valid, heteroskedasticity-robust estimates of the group-level ATE.

This is the **Sorted Group Average Treatment Effect (GATES)** approach from Chernozhukov et al. (2018).
:::

```{r}
#| label: gates-plot
#| fig-alt: "Error bar plot showing estimated ATE within each quintile of predicted CATEs, with overall ATE reference line"

gates.df <- data.frame(
  group = 1:num.rankings,
  ate = coef(ols),
  se = sqrt(diag(vcovHC(ols, "HC2")))
)
gates.df$lower <- gates.df$ate - 1.96 * gates.df$se
gates.df$upper <- gates.df$ate + 1.96 * gates.df$se

ggplot(gates.df, aes(x = factor(group), y = ate)) +
  geom_point(size = 3, color = "#13294B") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2,
                color = "#13294B") +
  geom_hline(yintercept = mean(tau.hat), linetype = "dashed",
             color = "#E07C3E") +
  labs(
    title = "Group Average Treatment Effects (GATES)",
    subtitle = "ATE within each quintile of predicted CATE",
    x = "CATE Quintile (1 = least benefit, 5 = most benefit)",
    y = "Estimated ATE (AIPW)"
  ) +
  annotate("text", x = 0.7, y = mean(tau.hat), vjust = -1,
           label = "Overall ATE", color = "#E07C3E", fontface = "italic")
```

### Step 2: CLAN (Characterization of Least and Most Affected Groups)

Who is in each group? Compare covariate means between the most-affected and least-affected quintiles.

```{r}
#| label: clan

top.group <- ranking == num.rankings
bottom.group <- ranking == 1

clan.results <- data.frame(
  variable = colnames(X),
  mean.bottom = sapply(X[bottom.group, ], mean),
  mean.top = sapply(X[top.group, ], mean)
)
clan.results$difference <- clan.results$mean.top - clan.results$mean.bottom

print(clan.results)
```

```{r}
#| label: clan-plot
#| fig-alt: "Horizontal bar chart showing difference in covariate means between top and bottom CATE quintiles"

ggplot(clan.results, aes(x = reorder(variable, abs(difference)),
                          y = difference)) +
  geom_col(fill = ifelse(clan.results$difference > 0, "#4B9CD3", "#E07C3E")) +
  coord_flip() +
  labs(
    title = "CLAN: How do the most-affected differ from least-affected?",
    subtitle = "Difference in covariate means: Top quintile - Bottom quintile",
    x = NULL, y = "Difference in means"
  )
```

::: callout-note
## Interpreting CLAN

CLAN tells you what **characterizes** the people who benefit most. Large bars mean that covariate strongly distinguishes high-benefit from low-benefit patients. You should see `comorbidities` and `age` dominating — the true effect modifiers.

This is **descriptive**, not causal. CLAN answers "who is in the top group?" not "would changing this covariate change the effect."
:::

### Step 3: RATE and TOC Curves

The **Rank-Average Treatment Effect (RATE)** evaluates how well our CATE predictions can be used for targeting. We use a split-sample approach for honest evaluation.

```{r}
#| label: rate-setup

# Split-sample approach for honest evaluation
train.idx <- sample(1:n, n / 2)
test.idx <- setdiff(1:n, train.idx)

# Train forest on training set
cf.train <- causal_forest(
  as.matrix(X[train.idx, ]), Y[train.idx], W[train.idx],
  W.hat = 0.5, num.trees = 2000, seed = 883
)

# Predict CATEs on test set (out-of-sample)
tau.hat.test <- predict(cf.train, as.matrix(X[test.idx, ]))$predictions

# Evaluation forest on test set (independent)
cf.eval <- causal_forest(
  as.matrix(X[test.idx, ]), Y[test.idx], W[test.idx],
  W.hat = 0.5, num.trees = 2000, seed = 884
)

# RATE: AUTOC (recommended default)
rate.result <- rank_average_treatment_effect(cf.eval, tau.hat.test)
print(rate.result)

# One-sided p-value
cat("\nOne-sided p-value:",
    pnorm(rate.result$estimate / rate.result$std.err,
          lower.tail = FALSE), "\n")
```

```{r}
#| label: toc-curve
#| fig-alt: "TOC curve showing treatment effect advantage as a function of the fraction treated, with highest predicted CATE treated first"

plot(rate.result, las = 1,
     main = "TOC Curve: Targeting by Predicted CATE",
     xlab = "Fraction treated (highest CATE first)",
     ylab = "Treatment effect advantage")
```

::: callout-tip
## How to Read the TOC Curve

The TOC (Targeting Operator Characteristic) curve shows what happens if you treat individuals **in order of predicted benefit**:

-   **X-axis:** Fraction of population treated (starting with highest predicted CATE)
-   **Y-axis:** Treatment effect for the treated fraction *minus* the overall ATE

If the curve is above zero on the left, your targeting rule successfully identifies high-benefit individuals. The **area under the curve (AUTOC)** summarizes overall targeting quality.

-   AUTOC \> 0 with p \< 0.05: targeting works better than random assignment
-   AUTOC ≈ 0: no useful heterogeneity for targeting (treat everyone or no one)
:::

::: callout-important
## Think About It #3

Consider the GATES plot, CLAN results, and RATE output together.

1.  CLAN shows you *who* is in the top group. Is that enough to build a targeting rule? What's missing?
2.  The RATE/AUTOC test tells you *whether* targeting works. But does it tell you the optimal *threshold* — how many people to treat?
3.  How might a budget constraint change which analysis (GATES, CLAN, RATE, or policy tree) is most useful?

**Preview:** Exercise 4 shows how `policytree` turns these insights into an explicit decision rule.
:::

------------------------------------------------------------------------

## Exercise 4: Policy Tree Preview

Given CATEs, can we learn a simple, interpretable treatment rule?

### Step 1: Extract Doubly Robust Scores

```{r}
#| label: dr-scores

# Doubly robust scores encode the causal information needed for policy learning
dr.scores <- double_robust_scores(cf)
head(dr.scores)

cat("Dimensions:", dim(dr.scores), "\n")
cat("Column 1 = E[Y(0)|X], Column 2 = E[Y(1)|X]\n")
```

### Step 2: Fit a Policy Tree

```{r}
#| label: policy-tree
#| fig-alt: "Decision tree showing optimal treatment assignment rule based on patient characteristics"

# Learn a shallow policy tree (depth = 2 for interpretability)
ptree <- policy_tree(as.matrix(X), dr.scores, depth = 2)

# Visualize the policy
plot(ptree, leaf.labels = c("Don't Treat", "Treat"))
```

::: callout-note
## What Is the Policy Tree Doing?

The policy tree learns a simple decision rule that maximizes expected welfare. It takes the DR scores (which encode the causal information) and finds the best depth-2 tree that assigns patients to treatment or control.

The result is an interpretable rule like: "Treat if comorbidities $\geq$ 2 AND age \> 55." This is much simpler than the full causal forest — it trades some optimality for interpretability.
:::

### Step 3: Evaluate the Policy

```{r}
#| label: policy-evaluation

# Who does the tree assign to treatment?
assignments <- predict(ptree, as.matrix(X))
cat("Fraction assigned to treatment:", mean(assignments == 2), "\n")
cat("Fraction assigned to control:  ", mean(assignments == 1), "\n")

# Compare ATE for treated vs not-treated groups
cat("\nTrue ATE for policy-tree-treated group:",
    round(mean(tau.true[assignments == 2]), 3), "\n")
cat("True ATE for policy-tree-untreated group:",
    round(mean(tau.true[assignments == 1]), 3), "\n")
cat("Overall true ATE:",
    round(mean(tau.true), 3), "\n")
```

::: callout-important
## Think About It #4

Compare the policy tree rule to the CLAN analysis:

1.  Does the policy tree use the same covariates that CLAN identified as important? Why or why not?
2.  A policy tree gives you an **explicit rule** (treat/don't treat). CLAN gives you **group descriptions** (who's in Q5 vs Q1). When would you prefer one over the other?
3.  This is a depth-2 tree. What happens if you increase depth? Is more complex always better?

**We'll formalize policy learning in Unit 4**, including welfare analysis, budget constraints, and optimal treatment rules.
:::

------------------------------------------------------------------------

## Wrap-Up

### Key Takeaways

1.  **Always test before interpreting.** CATE histograms *always* show spread — even when effects are constant. `test_calibration()` is the firewall between noise and signal.

2.  **The pipeline has a logical order.** Estimate → Test → Summarize → Evaluate → Decide. Each step answers a distinct question; skipping steps leads to overconfident claims.

3.  **Variable importance** $\neq$ BLP $\neq$ CLAN. Variable importance tells you *which* covariates drive heterogeneity (splitting frequency). BLP tells you *how* covariates relate to CATEs (linear projection). CLAN tells you *who* is in the top/bottom groups (covariate profiles). They're complementary, not redundant.

4.  **Split-sample evaluation is essential for RATE.** Using the same data to estimate CATEs and evaluate targeting quality is circular. The train/test split in the RATE analysis ensures honest evaluation.

5.  **Policy trees bridge estimation and action.** A CATE estimate tells you "this person's effect is -1.9." A policy tree tells you "treat everyone over 55 with 2+ comorbidities." The latter is what a program manager needs.

### Common Debugging Tips

| Problem | Likely Cause | Fix |
|-----------------------|-------------------------------|------------------|
| `causal_forest()` runs slowly | Too many trees or small `min.node.size` | Start with 2000 trees; increase if diagnostics are noisy |
| `test_calibration()` p \> 0.05 but you expected heterogeneity | Weak signal, small n, or noisy outcome | Increase n, reduce outcome noise, or try `tune.parameters = "all"` |
| Variable importance ranks noise variables high | Small n or high outcome noise | Focus on BLP significance, not raw importance |
| GATES confidence intervals all overlap | Forest can't distinguish groups precisely | This is a valid finding — heterogeneity may exist but isn't estimable at this sample size |
| `policy_tree()` error about scores | Wrong input format | Use `double_robust_scores(cf)`, not raw predictions |
| Different results on each run | Seed not set | Always set `set.seed()` before `causal_forest()` AND before fold creation |

### Connection to PS-3

Problem Set 3 asks you to apply these tools to a **different dataset** (Patient Engagement App, n=5,000) and extends the analysis in two important ways:

1.  **Meta-learners** (T-learner, X-learner) — alternative approaches to CATE estimation that don't use forests
2.  **Independent analysis** — you choose which diagnostics to run and how to summarize heterogeneity

The core `grf` pipeline from today is your foundation. PS-3 builds on it.

### Corrected Code-Along Reference

The instructor's corrected version of the Session 3.0 code-along is available here:

-   [Open in Google Colab](https://colab.research.google.com/drive/1N1RSV4LTA74N09QrwPTAtiolMxehqWbN)

------------------------------------------------------------------------

## Additional Resources

-   [GRF Documentation](https://grf-labs.github.io/grf/)
-   [GRF Diagnostics Vignette](https://grf-labs.github.io/grf/articles/diagnostics.html)
-   [GRF RATE Vignette](https://grf-labs.github.io/grf/articles/rate.html)
-   Athey, S., Tibshirani, J., & Wager, S. (2019). Generalized Random Forests. *Annals of Statistics*, 47(2), 1148-1178.
-   Chernozhukov, V. et al. (2018). Generic Machine Learning Inference on Heterogeneous Treatment Effects. *Econometrica*, 86(6).
-   Wager, S. (2024). *Causal Inference: A Statistical Learning Approach*. Chapter 6.