---
title: "Problem Set 7"
subtitle: "Advanced Topics (Optional)"
author: "HPM 883: Advanced Quantitative Methods"
date: "Due: Finals Week"
slug: "ps-7"
categories: [Problem Set, Advanced, Networks, IV, Transportability]
description: "Explore frontier methods in causal ML. Choose one of three advanced topics: network interference, instrumental variables + ML, or transportability."
format:
  html:
    toc: true
    toc-depth: 2
    code-tools: true
editor: visual
execute:
  eval: false
  echo: true
---

## Introduction

This optional problem set explores **frontier methods** beyond the core curriculum. It can replace any lower-scoring PS 1-6.

St. Null's Memorial Hospital faces methodological challenges that require advanced techniques. **Choose ONE** of the following three options.

---

## Option A: Network Interference

### Scenario

The Patient Engagement App shows **spillover effects** — patients influence each other within hospital units. Your task is to estimate treatment effects accounting for interference.

### Data

```{r}
#| label: load-option-a
#| eval: false

network_data <- read.csv("data/ps-7-network.csv")
```

- `Y`: Outcome
- `W`: Individual treatment
- `unit_id`: Hospital unit (cluster)
- `unit_treated_frac`: Fraction of unit receiving treatment
- Network adjacency matrix provided separately

### Tasks (100 points)

1. **Conceptual (20 points)**
   - Define direct and indirect (spillover) effects
   - Explain why SUTVA violations bias standard estimates
   - Describe the "exposure mapping" approach to interference

2. **Implementation (40 points)**
   - Estimate direct treatment effect using exposure-marginalized estimators
   - Estimate spillover effect (effect of peer treatment)
   - Compare to naive estimate ignoring interference

3. **Diagnostics (20 points)**
   - Test for presence of interference
   - Visualize how effects vary by unit treatment saturation

4. **Interpretation (20 points)**
   - What are the implications for program design?
   - Should the hospital treat individuals or whole units?

### Key References
- Aronow & Samii (2017) - Estimating average causal effects under interference
- Athey et al. (2018) - Exact p-values for network interference

---

## Option B: Instrumental Variables + ML

### Scenario

A natural experiment provides instrumental variable(s) for studying the effect of specialist care on outcomes. Distance to the nearest specialist hospital serves as an instrument.

### Data

```{r}
#| label: load-option-b
#| eval: false

iv_data <- read.csv("data/ps-7-iv.csv")
```

- `Y`: Health outcome
- `D`: Received specialist care (endogenous)
- `Z1`-`Z5`: Candidate instruments (distance, local capacity, etc.)
- `X1`-`X30`: Covariates

### Tasks (100 points)

1. **Conceptual (20 points)**
   - Explain LATE interpretation in the IV setting
   - Why might multiple weak instruments be problematic?
   - How does regularization help with many instruments?

2. **Implementation (40 points)**
   - Standard 2SLS with single best instrument
   - Lasso-IV with all candidate instruments
   - Compare estimates and standard errors

3. **Diagnostics (20 points)**
   - First-stage F-statistics
   - Weak instrument diagnostics
   - Over-identification tests (if applicable)

4. **Interpretation (20 points)**
   - Who are the compliers? How generalizable is LATE?
   - Policy recommendations given the estimates

### Key References
- Chernozhukov et al. Ch 11-12 - IV with many instruments
- Belloni et al. (2012) - Sparse models and LATE

---

## Option C: Transportability

### Scenario

The RCT was conducted at St. Null's main campus. Leadership wants to know if effects generalize to their rural affiliate network, which has different patient demographics.

### Data

```{r}
#| label: load-option-c
#| eval: false

# RCT data (source population)
rct_data <- read.csv("data/ps-7-transport-rct.csv")

# Target population survey
target_data <- read.csv("data/ps-7-transport-target.csv")
```

RCT has outcomes; target population has covariates only.

### Tasks (100 points)

1. **Conceptual (20 points)**
   - Define the PATE (population ATE) vs SATE (sample ATE)
   - Explain the transportability assumptions (S-ignorability)
   - When might effects not transport? Give examples.

2. **Implementation (40 points)**
   - Estimate SATE from RCT data
   - Estimate PATE using inverse probability of selection weighting (IPSW)
   - Use ML for the selection model

3. **Diagnostics (20 points)**
   - Compare covariate distributions between source and target
   - Assess overlap in selection scores
   - Sensitivity to selection model choice

4. **Interpretation (20 points)**
   - How different are SATE and PATE?
   - What covariates drive the difference?
   - Should the hospital expect similar effects at rural affiliates?

### Key References
- Dahabreh et al. (2020) - Extending inferences from RCTs
- Stuart et al. (2018) - Generalizability of randomized trial results
- Bareinboim & Pearl (2016) - Causal inference and data-fusion

---

## Submission Requirements

Push to your GitHub Classroom repository by **Finals Week deadline**:

1. Clearly indicate which option (A, B, or C) you completed
2. Rendered HTML and source .qmd file

---

## Evaluation Criteria

Each option is graded on the same scale:

| Component | Points |
|-----------|--------|
| Conceptual | 20 |
| Implementation | 40 |
| Diagnostics | 20 |
| Interpretation | 20 |

**Note:** This PS can replace your lowest-scoring PS 1-6 if the score is higher.
